Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed-particle hydrodynamics

### 2017-03-23 Manuscripts_GMD.tex
jonesm changes, mostly small, some larger comments flagged in red

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}
Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere.
Simulation of all possible hazards with one model is difficult due the fact that different physics of different length scales are dominant in different hazards. Our focus here is the hazards that volcanic ash poses to aircraft.

@@

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with finite volume method. A parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}.

Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that attempts to reproduce the entrainment process of eruption clouds with relatively simpler physics but with high order numerical accuracy and high spatial resolution.
A series of simulations based on SK-3D was reported, including determination of the entrainment coefficients of eruption columns as a function of height \citep{suzuki2010numerical}, the effect of wind field on entrainment coefficient \citep{suzuki20133d}, establishment of the relationship between the observable quantities of the eruption clouds and the eruption conditions at the vent \citep{suzuki2009three} (previously, such relationship is based on 1D models and greatly depends on empirical constants), investigation of the effect of the intensity of turbulence in the umbrella cloud on dispersion and sedimentation of tephra \citep{koyaguchi2009effect}.

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC developers have moved to include the effect of microphysics into the model while ATHAM developer have extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution.

Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N-phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adapting its unstructured finite volume solver.

To summarize, each 3D model has its own focus based on problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) are always a pair of conflict considerations in 3D plume simulations.\\

\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item Better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These are enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term is treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases only leads to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plume are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenario. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and comuptationaly efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

Open source availability and  the easy extension features of SPH facilitate development of a more comprehensive community driven model.

\subsection{Our contributions}
SPH is still a relatively new method in computational fluid dynamics compared to mesh-based methods, like Finite Volume. Implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling.

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adapted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

\section{Physical model} \label{sec:physics-model}
\subsection{Description of the model}
During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising up to several kilometers to tens of kilometers until its heat exhausted. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. During the process of plume rising up, relatively larger particles might separate from main stream of the plume, falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor condenses to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor  further heats up entrained air and make the bulk density further dilute. The entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation.

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases does not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term does not show up with this assumption).

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors would play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refered as weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano on April 4th 1982 \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcanic plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

\subsection{Governing equations}
Based above assumptions, the governing equation of our model is given as following, which is the same as governing equations of SK-3D \citep{suzuki2005numerical}.
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot (\rho \textbf{v}) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot (\rho \xi \textbf{v}) = 0 \label{eq:gov-cs-ks}\\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot (\rho \textbf{v} \textbf{v} + p\textbf{I}) = \rho \textbf{g} \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot [(\rho E + p )\textbf{v}] = \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{align}
$\xi$ is the mass fraction of ejected material.
$E = e + K $ is total energy which is a summation of kinetic energy $K$ and internal energy $e$.
An additional equation is required to close the system. In this model, the equation for closing the system is the EOS (equation of state) for ideal gas.
\begin{equation}
p = (\gamma_m - 1)\rho e \label{eq:EOS}
\end{equation}
Where 
\begin{equation}
\gamma_m = R_m/C_{vm} + 1 \label{eq:gov-gm}
\end{equation}
\begin{equation}
R_m = n_gR_g + n_aR_a  \label{eq:gov-Rm}
\end{equation}
\begin{equation}
C_{vm} = n_s C_{vs} + n_g C_{vg} + n_a C_{va} \label{eq:gov-Cvm}
\end{equation}
\begin{equation}
n_a = 1 - \xi \label{eq:gov-na}
\end{equation}
\begin{equation}
n_g = \xi n_{g0} \label{eq:gov-ng}
\end{equation}
\begin{equation}
n_s = \xi - n_g \label{eq:gov-ns}
\end{equation}
Where, $C_v$ is specific heat with constant volume, $n$ is mass fraction, $R$ is gas constant. The subscript 
$m$ represents mixture of ejected material and air, $s$ represents solid portion in ejected material, $g$ represents gas portion in the ejected material and $a$ represents air.

@@

\subsubsection{Velocity inlet}
At the vent, temperature of erupted material $T$, eruption velocity $\textbf{v}$, the mass fraction of vapor in erupted material $n_{g0}$ and mass discharge rate $\dot M$ is given. The pressure of erupted material $p$ is assumed to be the same as ambient pressure for pressure-balanced eruption. The radius of vent is determined from $\rho$, $\dot M$ and $\textbf{v}$. Equation (\ref{eq:erupt_bc_rho}) ~ (\ref{eq:erupt_bc_e}) gives velocity outlet boundary condition wrote in terms of primitive variables.
\begin{align}
\rho =const = p/(R_m T) \label{eq:erupt_bc_rho} \\
\xi=const=1 \label{eq:erupt_bc_xi}\\
\textbf{v} = const =\{u,v,w\}^T \label{eq:erupt_bc_v}\\
\dfrac{\partial e}{\partial n}=\dot M e /(\pi r^2) \label{eq:erupt_bc_e}
\end{align} 
Where $r$ is the radius of the vent.

@@

\subsection{Weighting function}
As described in previous section, the weighting function (or kernel function) is used to replace Dirac function and should satisfy Eq.  (\ref{eq:SPH_kernel_delta}). So it can be viewed as approximation form of Dirac function and satisfy normalization condition:
\begin{equation}
\int	 w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime = 1
\label{eq:SPH-kernel-normalization-prop}
\end{equation}
Besides normalization, weighting function of particle $a$ has to be symmetric with respect to $a$ to ensure that neighbor particles locate at the same distance away from $a$ contribute equally to SPH summation equation, see Eq. (\ref{eq:SPH-kernel-symmetric}). Weighting function also needs to satisfy conditions such as positivity and compact support. In addition, kernel function must be monotonically decreasing with the distance between particles.\\
\begin{equation}
w(\textbf{x}- \textbf{x} \prime, h) = w(\textbf{x} \prime - \textbf{x}, h)
\label{eq:SPH-kernel-symmetric}
\end{equation}

>>>>

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}
Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere.
Simulation of all possible hazards with one model is difficult due to the fact that different length scales dominate in different hazards. Our focus here is the hazard that volcanic ash poses to aircraft.

@@

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with a finite volume method. A parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}.

Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that attempts to reproduce the entrainment process of eruption clouds with relatively simple physics but with high order numerical accuracy and high spatial resolution.
A series of simulations based on SK-3D was reported, including determination of the entrainment coefficients of eruption columns as a function of height \citep{suzuki2010numerical}, the effect of wind field on entrainment coefficient \citep{suzuki20133d}, establishment of the relationship between the observable quantities of the eruption clouds and the eruption conditions at the vent \citep{suzuki2009three} (previously, such relationship is based on 1D models and greatly depends on empirical constants), and investigation of the effect of the intensity of turbulence in the umbrella cloud on dispersion and sedimentation of tephra \citep{koyaguchi2009effect}.

While SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM (Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, \textcolor{red}{immediately dynamics} and thermodynamic equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC development has begun to include the effect of microphysics into the model while ATHAM development has extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution.

Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N-phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based on the open source numerical solver OpenFOAM \textcolor{red}{cite?}, adapting its unstructured finite volume solver.

To summarize, each 3D model has its own focus based on the problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depending on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depending on number of governing equations, resolution, numerical methods and parallel techniques) are always a pair of conflicting considerations in 3D plume simulations.\\

\subsection{Features of SPH}
To the bext of our knowledge, all of the existing 3D plume models are mesh based Eulerian methods, and there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item Better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These are enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term is treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases only leads to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and comuptationaly efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

Open source availability and the relatively easy extensibility of SPH will facilitate development of a more comprehensive community driven model.

\subsection{Our contributions}
SPH is still a relatively new method in computational fluid dynamics compared to mesh-based methods, like Finite Volume. Implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in the simulation of free surface flow, such as breaking-waves and floods. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling.

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adapted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

\section{Physical model} \label{sec:physics-model}
\subsection{Description of the model}
During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising up to several kilometers to tens of kilometers until its heat is exhausted. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. During the process of plume rising up, relatively larger particles might separate from main stream of the plume, falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor condenses to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor  further heats up entrained air and further dilutes the bulk density. The entrained vapor might also experience a similar process and impact plume development. Particle aggregation processes \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation.

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamic equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases does not show up under this assumption). Immediate dynamic equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term does not show up with this assumption).

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors would play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not yet considered in this model. Introducing wind effects in our model requires dynamic pressure boundary conditions, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refered as a weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano on April 4th 1982 \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcanic plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

\subsection{Governing equations}
Based above assumptions, the governing equation of our model is given as following, which is the same as the governing equations of SK-3D \citep{suzuki2005numerical}.
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot (\rho \textbf{v}) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot (\rho \xi \textbf{v}) = 0 \label{eq:gov-cs-ks}\\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot (\rho \textbf{v} \textbf{v} + p\textbf{I}) = \rho \textbf{g} \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot [(\rho E + p )\textbf{v}] = \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{align}
$\xi$ is the mass fraction of ejected material.
$E = e + K $ is total energy which is a summation of kinetic energy $K$ and internal energy $e$.
\textcolor{red}{should define all variates when they first appear, even the ``obvious'' ones.}
An additional equation is required to close the system. In this model, the equation for closing the system is the EOS (equation of state) for ideal gas.
\begin{equation}
p = (\gamma_m - 1)\rho e \label{eq:EOS}
\end{equation}
Where 
\begin{equation}
\gamma_m = R_m/C_{vm} + 1 \label{eq:gov-gm}
\end{equation}
\begin{equation}
R_m = n_gR_g + n_aR_a  \label{eq:gov-Rm}
\end{equation}
\begin{equation}
C_{vm} = n_s C_{vs} + n_g C_{vg} + n_a C_{va} \label{eq:gov-Cvm}
\end{equation}
\begin{equation}
n_a = 1 - \xi \label{eq:gov-na}
\end{equation}
\begin{equation}
n_g = \xi n_{g0} \label{eq:gov-ng}
\end{equation}
\begin{equation}
n_s = \xi - n_g \label{eq:gov-ns}
\end{equation}
Where, $C_v$ is specific heat with constant volume, $n$ is mass fraction, $R$ is gas constant. The subscript 
$m$ represents mixture of ejected material and air, $s$ represents solid portion in ejected material, $g$ represents gas portion in the ejected material and $a$ represents air.

@@

\subsubsection{Velocity inlet}
At the vent, temperature of erupted material $T$, eruption velocity $\textbf{v}$, the mass fraction of vapor in erupted material $n_{g0}$ and mass discharge rate $\dot M$ are given. The pressure of erupted material $p$ is assumed to be the same as ambient pressure for pressure-balanced eruption. The radius of vent is determined from $\rho$, $\dot M$ and $\textbf{v}$. Equation (\ref{eq:erupt_bc_rho}) ~ (\ref{eq:erupt_bc_e}) gives velocity outlet boundary condition wrote in terms of primitive variables.
\begin{align}
\rho =const = p/(R_m T) \label{eq:erupt_bc_rho} \\
\xi=const=1 \label{eq:erupt_bc_xi}\\
\textbf{v} = const =\{u,v,w\}^T \label{eq:erupt_bc_v}\\
\dfrac{\partial e}{\partial n}=\dot M e /(\pi r^2) \label{eq:erupt_bc_e}
\end{align} 
Where $r$ is the radius of the vent.

@@

\subsection{Weighting function}
As described in the previous section, the weighting function (or kernel function) is used to replace the Dirac function and should satisfy Eq.  (\ref{eq:SPH_kernel_delta}). So it can be viewed as an approximate form of the Dirac function and satisfy the normalization condition:
\begin{equation}
\int	 w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime = 1
\label{eq:SPH-kernel-normalization-prop}
\end{equation}
Besides normalization, the weighting function of particle $a$ has to be symmetric with respect to $a$ to ensure that neighbor particles located at the same distance away from $a$ contribute equally to SPH summation equation, see Eq. (\ref{eq:SPH-kernel-symmetric}). The weighting function also needs to satisfy conditions such as positivity and compact support. In addition, the kernel function must be monotonically decreasing with the distance between particles.\\
\begin{equation}
w(\textbf{x}- \textbf{x} \prime, h) = w(\textbf{x} \prime - \textbf{x}, h)
\label{eq:SPH-kernel-symmetric}
\end{equation}

There is a wide variety of possible weighting functions, such as spline functions (with different orders) and Gaussian functions. Generally, the accuracy increases with the order of the polynomials of the kernel function, but the computational time also increases as number of interactions increase. 
We are adopting a truncated Gaussian function as the weighting function in our simulation.
\begin{equation}
w(\textbf{x} - \textbf{x}_b) = 
\begin{cases} 
      \dfrac{1}{(h \sqrt{\pi})^d} exp \left[- \left(\dfrac{\textbf{x} - \textbf{x} \prime}{h}\right)^2 \right] &  \vert \textbf{x} - \textbf{x} \prime \vert \leq 3h\\
      0 & \text{Otherwise}
\end{cases}
\label{eq:SPH-kernel}
\end{equation}
Where $d$ is number of dimensions.
The derivative of the weighting function:
\begin{equation}
\nabla w(\textbf{x} - \textbf{x}_b) = 
\begin{cases} 
      -2\left(\dfrac{\textbf{x} - \textbf{x} \prime}{h^2}\right) \dfrac{1}{(h \sqrt{\pi})^d} exp \left[- \left(\dfrac{\textbf{x} - \textbf{x} \prime}{h}\right)^2 \right] &  \vert \textbf{x} - \textbf{x} \prime \vert \leq 3h\\
      0 & \text{Otherwise}
\end{cases}
\end{equation}
